---
title: "tidyverse"
author: "David John Baker"
date: "12/23/2018"
output:
  pdf_document: default
  html_document: default
---

## Lesson Goals

* Be able to explain tidy data
* Explain the five tidyverse verbs
* Perform basic indexing
* Import and Export data from R 

## tidyverse + tidydata

One of the most important concepts data science and R is the idea of [tidydata](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).

The idea behind tidy data is that...

1. Each variable forms a column
2. Each observation forms a row.
3. Each type of observation unit forms a table.

If your data is in this format, then you can do almost anything with the tidyverse.

In order to use the tidyverse, you first need to install it.
```{r}
# install.packages("tidyverse") # Only need to do this once! 
library(tidyverse)
```

## Five Verbs

The five tidyverse verbs come from the dplyr package.
More information on this package can be found [here](https://dplyr.tidyverse.org/) along with these descriptions.

* ```mutate()``` adds new variables that are functions of existing variables
* ```select()``` picks variables based on their names.
* ```filter()``` picks cases based on their values.
* ```summarise()``` reduces multiple values down to a single summary.
* ```arrange()``` changes the ordering of the rows.

We can think the verbs as happening in the logical order you would want to grab them.
Each of the verbs is also going to be connected to one another with the pipe operator.
The idea behind the pipe or ```%>%`` is that the output of the last line is the first argument of the new function.

For example, if we wanted to make a small table that only had data from El Paso from 2011, then only get the first and fifth columns we would run the following code:

```{r}
str(txhousing)
txhousing_only_el_paso <- txhousing[txhousing$city == "El Paso",]
iris_only_only_el_paso_2005_2011 <- txhousing_only_el_paso[txhousing_only_el_paso$year >= 2011,]
iris_only_only_el_paso_2005_2011[,c(1,5)]
```
Which is a bit verbose.

In order to do this with the tidyverse, you would start with the dataset, the run two verbs over it, connected with the pipe.

```{r}
library(tidyverse)
iris_tibble <- as.tibble(iris)

txhousing %>%
  filter(city == "El Paso") %>%
  filter(year >= 2011) %>%
  select(1,5)

```

Both create the same output, but one is much easier to read.

We will now explore a dataset using the five verbs in the dplyr package.
You use each of the five verbs as you would in English to think about how you want to manipulate your data.

They key to using the tidyverse is the %>% operator (the pipe operator).
It works by taking output from what is before it and piping it to the next command.

## Economics Data

The dataset here comes from housing sales data in Texas provided by the TAMU real estate centre.

Variable | Description 
---------|-----------------
city | Name of MLS area
year,month,date |Date
sales | Number of sales
volume | Total value of sales
median | Median sale price
listings | Total active listings
inventory | "Months inventory" : amount of time it would take to sell all current listings at current pace of sales.

### Select

The select command works by "Selecting" the columns you wish to work with.
It can either take the index of the column using numbers, or the text.
There are other options like asking for columns that start or end with certain text.

```{r}
txhousing %>%
  select(1,2)

txhousing %>%
  select(1:3)

txhousing %>%
  select(city, sales:median)

txhousing %>%
  select(starts_with("ci"))

txhousing %>%
  select(-city)
```


### Filter

Once we have the columns we want to work with, we can then pick the rows that are of interest.
We do this with the filter function.
Here when asking for matches of character strings, you need to use the ``==``.
R will remind you if you forget.
The filter command can be combined with the logical operators. 
Remember this includes negation operators. 

```{r}
txhousing %>%
  filter(city == "El Paso")

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio")

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) 

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  filter(month != 1)
```

### Mutate

The mutate command will create new variables.

```{r}
txhousing %>%
  mutate(zSales = scale(sales))

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  filter(month != 1) %>%
  mutate(zScale = scale(sales))
```


### Arrange

Arrange will sort our data. 

```{r}
txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  filter(month != 1) %>%
  mutate(zScale = scale(sales)) %>%
  arrange(sales)

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  filter(month != 1) %>%
  mutate(zScale = scale(sales)) %>%
  arrange(sales, -year)

```

### Group By and Sumemrise 

Often we also want to perform the same type of calculation on a group in our dataset.
For this we need to group our data, then use the summarize command. 
We can also use the ``n()`` function to count the number of observations in each group.

```{r}
txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  group_by(year) %>%
  summarise(mean = mean(sales))
  

txhousing %>%
  filter(city == "El Paso" | city == "San Antonio") %>%
  filter(year >= 2004) %>%
  group_by(year) %>%
  summarise(mean = mean(sales), n = n())
```
