---
author: "David John Baker"
date: "04/05/2020"
output:
  pdf_document: default
  html_document: default
---

Let's now continue to explore the tidyverse with the `dplyr` package.
Here we'll learn about five verbs and the pipe operator.

```{r}
library(dplyr) # for manipulating data

library(readr) # for getting data
library(ggplot2) # for plotting data

tips <- read_csv("tips.csv")

# You can change where this is output above in "Settings (by knit) > Chunk Output in Console"
tips

```

The first verb we will use is `select()` which lets us choose colunmns. 

```{r}
select(tips, total_bill)

select(tips, -X1)

select(tips, tip:size)

select(tips, starts_with("s"))
```

Now let's check out filter!

```{r}

filter(tips, day == "Sun")

filter(tips, tip > 5)

filter(tips, sex == "Male" & smoker == "Yes")

filter(tips, sex == "Male" | smoker == "Yes")

```

Let's now return to our example from before to see `mutate()` in action. 
It's the same output, but right now looks a bit different since it's simplified.

```{r}
mutate(tips, gbp_total = total_bill * 0.81)

# R doesn't care about spacing!! 
mutate(tips, 
       gbp_total = total_bill * 0.81,
       gbp_tip = tip * 0.81)

```

Up until this point, we have been doing some very basic commands.
What if we want to get super fancy?
We could save a bunch of intermediate objects into memory... or we could use the %>% operator! 

```{r}
select(tips, total_bill)

tips %>%  # and then! 
  select(total_bill)

```

We can then build up more complex commands.

```{r}
tips %>%
  select(total_bill, tip, sex, smoker) %>%
  filter(sex == "Male" & smoker == "Yes") %>%
  mutate(gbp_total_bill = total_bill * 0.81,
         gbp_tip = tip * 0.81)

```

But what if we wanted to investigate differences between smoker and non smokers for tipping?
For that we need group_by and summerise! 

```{r}
tips %>%
  group_by(smoker)

tips %>%
  group_by(smoker) %>%
  summarise(mean = mean(tip),
            count = n())

```

Lastly let's arrange our output here so that those who tip more are on top! 

```{r}
tips %>%
  group_by(smoker) %>%
  summarise(mean = mean(tip),
            count = n()) %>%
  arrange(desc(mean))
```

